\vspace{-15mm}\fontsize{21pt}{10pt}\selectfont\textbf{Wordy Manuscript Title: \\Even Wordier Manuscript Subtitle\thanks{Draft manuscript. Please do not cite without the authors' permission.}}

### 2022-09-01 sections/discussion.tex
Change: Update manuscript based on reviewer comments

  Futher benchmarking of co-occurrence networks could be pursued through the utilization of literature-based interactions~\cite{lima-mendezDeterminantsCommunityStructure2015a} or biological benchmark interaction data~\cite{sungGlobalMetabolicInteraction2017a}.
  Furthermore, \ac{micone} could be extended to enable the processing of metagenomics sequencing data, facilitating the analysis of a much larger and diverse range of datasets and domains of life.

>>>>

  Further, benchmarking of co-occurrence networks could also be pursued through the utilization of literature-based interactions~\cite{lima-mendezDeterminantsCommunityStructure2015a} or biological benchmark interaction data~\cite{sungGlobalMetabolicInteraction2017a}.
  However, direct comparisons between co-occurrence networks and directly measured interactions are difficult to interpret and highly debated~\cite{hiranoDifficultyInferringMicrobial2019,gobernaCautionaryNotesUse2022}.
  Additionally, \ac{micone} could be extended to enable the processing of metagenomics sequencing data, facilitating the analysis of a much larger and diverse range of datasets and domains of life.

### 2022-09-01 sections/methods.tex
Fix: Changes based on reviewer comments
Change: Update manuscript based on reviewer comments

  \subsection*{OTU and ESV Processing (OP)}
  \vspace{-5mm}
  This module deals with normalization, filtering, forking, grouping and applying transformations to the \ac{otu} or \ac{esv} counts matrix.
  Normalization of the count matrix involves converting the count matrix of read counts into a count matrix containing relative abundances.
  The module also supports rarefaction, which is a normalization technique used to overcome the bias that might arise due to variable sampling depth in different samples.
  This is performed either by sub-sampling or by normalization of the matrix to the lowest sampling depth~\cite{Weiss2015}.
  Although the pipeline supports normalization and rarefaction, these modules are turned off by default and the analysis reported in the paper do not normalize or rareify the counts matrices.
  However, most of the network inference methods perform normalization and other transformation operations on the counts matrix as a part of their workflow.
  Filtering, is performed to remove samples or features (\ac{otu}s or \ac{esv}s) from the counts matrix that are sparse.
  By default, when the OP module is ``on'', the samples are filtered out if the total reads in a sample are less than 500 and features are filtered out if the relative abundance is less than 1\%, prevalence (percentage of samples containing feature) is less than 5\% and count sum across all the samples is less than 100.
  When the OP module is ``off'', the filtering is still performed but only to remove those features that are very low in presence.
  The parameters used are given in Table~\ref{tab:micone_tools}.
  The forking operation splits the count matrix into multiple matrices based on sample metadata column, this is useful for example to compare case vs. control.
  The group operation transforms the \ac{otu} or \ac{esv} count matrix into a taxonomic count matrix at the requested level by adding up counts that map to the same taxonomy and is carried out at the end of the OP step.
  Finally, transformations are performed in order to correct for and overcome the compositional bias that is inherent in the counts matrix (in the analysis performed in the study these are handled by the network inference algorithm).
  All the modules in this step were implemented using functions from the \href{http://biom-format.org/}{biom-format} Python package~\cite{mcdonaldBiologicalObservationMatrix2012}.

@@

  \subsubsection*{Bootstrapping and p-value calculation}
  \vspace{-5mm}
  For all correlation-based methods $k < w$, $1000$ permutations of the original \ac{otu} counts data were generated~\cite{Watts2018}.
  All the direct association based methods used in the study have their own regularization methods built-in and hence do not need to undergo bootstrapping.
  The correlations in the permuted \ac{otu} tables are recalculated using the different correlation-based algorithms.
  Finally, the p-value is determined based on how often a more extreme association is observed for randomly permuted data.

@@

  \subsection*{Network Variability}

>>>>

  \subsection*{OTU and ESV Processing (OP)}
  \vspace{-5mm}
  This module deals with normalization, filtering, forking, grouping and applying transformations to the \ac{otu} or \ac{esv} counts matrix.
  Normalization of the count matrix involves converting the count matrix of read counts into a count matrix containing relative abundances.
  The module also supports rarefaction, which is a normalization technique used to overcome the bias that might arise due to variable sampling depth in different samples.
  However, due to the potential biases and false positives~\cite{mcmurdieWasteNotWant2014,chaoCoveragebasedRarefactionExtrapolation2012} that might arise during the process, the rarefaction module is disabled by default and can be enabled in the configuration if needed.
  This is performed either by sub-sampling or by normalization of the matrix to the lowest sampling depth~\cite{Weiss2015}.
  Although the pipeline supports normalization and rarefaction, these modules are turned off by default and the analysis reported in the paper do not normalize or rareify the counts matrices.
  However, most of the network inference methods perform normalization and other transformation operations on the counts matrix as a part of their workflow.
  Filtering, is performed to remove samples or features (\ac{otu}s or \ac{esv}s) from the counts matrix that are sparse.
  By default, when the OP module is ``on'', the samples are filtered out if the total reads in a sample are less than 500 and features are filtered out if the relative abundance is less than 1\%, prevalence (percentage of samples containing feature) is less than 5\% and count sum across all the samples is less than 100.
  When the OP module is ``off'', the filtering is still performed but only to remove those features that are very low in presence.
  The parameters used are given in Table~\ref{tab:micone_tools}.
  The forking operation splits the count matrix into multiple matrices based on sample metadata column, this is useful for example to compare case vs. control.
  The group operation transforms the \ac{otu} or \ac{esv} count matrix into a taxonomic count matrix at the requested level by adding up counts that map to the same taxonomy and is carried out at the end of the OP step.
  Finally, transformations are performed in order to correct for and overcome the compositional bias that is inherent in the counts matrix (in the analysis performed in the study these are handled by the network inference algorithm).
  All the modules in this step were implemented using functions from the \href{http://biom-format.org/}{biom-format} Python package~\cite{mcdonaldBiologicalObservationMatrix2012}.

@@

  \subsubsection*{Bootstrapping and p-value calculation}
  \vspace{-5mm}
  For all correlation-based methods $k \le w$, $1000$ permutations of the original \ac{otu} counts data were generated~\cite{Watts2018}.
  All the direct association based methods used in the study have their own regularization methods built-in and hence do not need to undergo bootstrapping.
  The correlations in the permuted \ac{otu} tables are recalculated using the different correlation-based algorithms.
  Finally, the p-value is determined based on how often a more extreme association is observed for randomly permuted data.

@@

  \subsection*{Network variability}

### 2022-09-01 sections/results.tex
Change: Update manuscript based on reviewer comments

  Since no database performs better than others against mock datasets, the choice of which database to use could be driven by other reasons.
  One user-specific way to choose, would be based on the known representation of taxa for the microbiome of interest (see Discussion).
  Another reason could be the frequency of updates and the potential for future growth.
  Since the \ac{gg} database is used frequently as a 16S reference database in studies~\cite{Balvociute2017}, to enable easier comparison with published networks in literature, we set \ac{gg} as the default for taxonomy assignment.
  On the other hand, since the degree of similarity of assignments between \ac{ncbi} and \ac{gg} is fairly high (Figure~\ref{fig:figure_s5}), and due to its regular revision and maintenance the \ac{ncbi} database is also a good choice for the user.
  The TA step results in a taxonomic counts table that is used as input to the subsequent steps of the pipeline.
  The count tables at different levels can be obtained by aggregation, for example, Genus count tables were obtained by summing up the counts of the lower taxonomy levels (Species and \ac{otu}) that map to the same higher taxonomy level name.

@@

  Similar to what was done for the previous steps of the pipeline, we evaluated the network inference algorithms and the final consensus network using synthetic interaction data.
  For this purpose, we generated synthetic interaction data using the ``NorTA''~\cite{Kurtz2015} and ``seqtime''~\cite{faustSignaturesEcologicalProcesses2018} methods (see Methods), which have been previously used for benchmarking network inference methods.
  For each method, an \ac{otu} counts table is generated based on the selected interaction network topology and properties.
  This counts table is used as the input to the \ac{micone} pipeline to generate predicted associations.
  The interaction network used to generate the counts table is used as the source of true interactions to calculate the precision and sensitivity for each network inference algorithm for the NorTA datasets (Figure \ref{fig:figure5}).
  The goal of the consensus network algorithms is to increase the precision of the resultant consensus network by removing edges that are either not present in most of the inference methods or whose association strength is low.
  In Figure \ref{fig:figure5} we show that the consensus algorithm, especially the scaled-sum method, captures true associations with more precision, but as a consequence the sensitivity of the predictions is reduced.
  Overall, we see that the scaled-sum method for $p=1.0$ performs the best in terms of average precision ($1.000$).
  However, the scaled-sum method for $p=0.667$ has the second-highest average precision ($0.985$) and does not underperform in terms of sensitivity.
  Among the individual inference methods, we observe that \ac{spieceasi} has the best average precision ($0.944$)
  Therefore, the consensus networks provide the means to obtain a short list of associations that would have a high probability of being present in the real interaction network.
  Furthermore, we also find that the consensus algorithm performs drastically better in terms of precision (0.820 for scaled-sum method vs. 0.624 for \ac{spieceasi}) in recapturing the original interactions in the ``seqtime'' data.
  The precision and sensitivity graphs on the seqtime data can be found in the Supplementary section (Figure~\ref{fig:figure_s7}).

>>>>

  Since no database performs better than others against mock datasets, the choice of which database to use could be driven by other reasons.
  One user-specific way to choose, would be based on the known representation of taxa for the microbiome of interest (see Discussion).
  Another reason could be the frequency of updates and the potential for future growth.
  Since the \ac{gg} database is used frequently as a 16S reference database in studies~\cite{Balvociute2017}, to enable easier comparison with published networks in literature, we set \ac{gg} as the default for taxonomy assignment.
  On the other hand, since the degree of similarity of assignments between \ac{ncbi} and \ac{gg} is fairly high (Figure~\ref{fig:figure_s5}), and due to its regular revision and maintenance the \ac{ncbi} database is also a good choice for the user.
  The TA step results in a taxonomic counts table that is used as input to the subsequent steps of the pipeline.
  The count tables at different levels can be obtained through aggregation, for example, Genus count tables were obtained by summing up the counts of the lower taxonomy levels (Species and \ac{otu}) that map to the same higher taxonomy level name.

@@

  Similar to what was done for the previous steps of the pipeline, we evaluated the network inference algorithms and the final consensus network using synthetic interaction data.
  For this purpose, we generated synthetic interaction data using the ``NorTA''~\cite{Kurtz2015} and ``seqtime''~\cite{faustSignaturesEcologicalProcesses2018} methods (see Methods), which have been previously used for benchmarking network inference methods~\cite{Kurtz2015,Weiss2016}.
  For each method, an \ac{otu} counts table is generated based on the selected interaction network topology and properties.
  This counts table is used as the input to the \ac{micone} pipeline to generate predicted associations.
  The interaction network used to generate the counts table is used as the source of true interactions to calculate the precision and sensitivity for each network inference algorithm for the NorTA datasets (Figure \ref{fig:figure5}).
  The goal of the consensus network algorithms is to increase the precision of the resultant consensus network by removing edges that are either not present in most of the inference methods or whose association strength is low.
  In Figure \ref{fig:figure5} we show that the consensus algorithm, especially the scaled-sum method, captures true associations with more precision, but as a consequence the sensitivity of the predictions is reduced.
  Overall, we see that the scaled-sum method for $p=1.0$ performs the best in terms of average precision ($1.000$).
  However, the scaled-sum method for $p=0.667$ has the second-highest average precision ($0.985$) and does not underperform in terms of sensitivity.
  Among the individual inference methods, we observe that \ac{spieceasi} has the best average precision ($0.944$)
  Therefore, the consensus networks provide the means to obtain a short list of associations that would have a high probability of being present in the real interaction network.
  Furthermore, we also find that the consensus algorithm performs drastically better in terms of precision (0.820 for scaled-sum method vs. 0.624 for \ac{spieceasi}) in recapturing the original interactions in the ``seqtime'' data.
  The precision and sensitivity graphs on the seqtime data can be found in the Supplementary section (Figure~\ref{fig:figure_s7}).

### 2022-09-01 sections/supplementary.tex
Fix: Changes based on reviewer comments
Change: Update manuscript based on reviewer comments

@@

  The choice of taxonomy database was found to be the most important factor in the inference of microbial co-occurrence networks, contributing $65.4\%$ of the total variance.
  The frequent changes in the taxonomy nomenclature coupled with the frequency of updates to the various 16S reference databases create inherent differences \cite{Balvociute2017} in taxonomy hierarchies in these databases.
  Our analysis revealed that no particular reference database performs better than the others across the different mock dataset benchmarks.
  The default reference database in the pipeline is the \ac{gg} reference database along with the ``Naive Bayes'' classifier as the query tool.
  The reason for our choice stems from the popularity of the \ac{gg} database~\cite{parkEvaluation16SRRNA2018} in taxonomic studies, which would enable easy comparison across datasets.
  However, for newer studies we recommend using SILVA database because of its size and taxonomic comprehensiveness~\cite{iiRESCRIPtReproducibleSequence2021} and since \ac{gg} has not been updated since 2013.
  Additionally, we also enable users to use custom databases~\cite{Ritari2015,iiRESCRIPtReproducibleSequence2021} with the BLAST and Naive Bayes classifiers that are incorporated into the pipeline (from \ac{qiime2}).
  We suggest that that choice of the database should be made based on possible reported or inferred biases in the representation of given biomes in a specific databases~\cite{Balvociute2017,iiRESCRIPtReproducibleSequence2021}, as choosing taxon-specific databases have also been observed to compromise classification~\cite{rmarcelinoUseTaxonspecificReference2020}.

>>>>

  {\color{red} Tables S1-S5 will be moved to GitHub}

@@

  The choice of taxonomy database was found to be the most important factor in the inference of microbial co-occurrence networks, contributing $65.4\%$ of the total variance.
  The frequent changes in the taxonomy nomenclature coupled with the frequency of updates to the various 16S reference databases create inherent differences \cite{Balvociute2017} in taxonomy hierarchies in these databases.
  Our analysis revealed that no particular reference database performs better than the others across the different mock dataset benchmarks.
  The default reference database in the pipeline is the \ac{gg} reference database along with the ``Naive Bayes'' classifier as the query tool.
  The reason for our choice stems from the popularity of the \ac{gg} database~\cite{parkEvaluation16SRRNA2018} in taxonomic studies, which would enable easy comparison across datasets.
  However, for newer studies we recommend using SILVA database because of its size and taxonomic comprehensiveness~\cite{iiRESCRIPtReproducibleSequence2021} and since \ac{gg} has not been updated since 2013.
  Additionally, a particular database might be more appropriate than the rest based on specific requirements.
  For example, in order to generate a dataset that is compatible with the \ac{mind} platform~\cite{huResourceComparisonIntegration2022} \ac{ncbi} is the most appropriate choice as it guarantees compatibility of taxonomic hierarchy and therefore comparability with other datasets.
  Furthermore, we also enable users to use custom databases~\cite{Ritari2015,iiRESCRIPtReproducibleSequence2021} with the BLAST and Naive Bayes classifiers that are incorporated into the pipeline (from \ac{qiime2}).
  We suggest that that choice of the database should be made based on possible reported or inferred biases in the representation of given biomes in a specific databases~\cite{Balvociute2017,iiRESCRIPtReproducibleSequence2021}, as choosing taxon-specific databases have also been observed to compromise classification~\cite{rmarcelinoUseTaxonspecificReference2020}.
